---
title: "Tutorial for main functions in longsurr package"
output:
  html_document:
---

```{r global_options, echo=FALSE}
knitr::opts_chunk$set(fig.width=13, warning=FALSE, tidy=TRUE, tidy.opts=list(width.cutoff=60))
```

In this tutorial, we will go through examples using the functions in the **longsurr** package. First, install the package from CRAN (or Github) and load it.

```{r results = "hide", message=FALSE}
#from CRAN:
#install.packages("longsurr")
#or Github:
#install.packages("devtools")  
#library(devtools)
#devtools::install_github("laylaparast/longsurr")
library(longsurr)

```

This package offers methods to (1) estimate the proportion of treatment effect on a primary outcome that is explained by the treatment effect on a longitudinal surrogate marker i.e., for surrogate marker evaluation, and (2) to directly estimate the treatment effect on a longitudinal surrogate marker. All methods are described in detail in Agniel and Parast (2021). Evaluation of Longitudinal Surrogate Markers, Biometrics, 77(2): 477-489, (<https://doi.org/10.1111/biom.13310>) and Wang et al (2025) "Semiparametric Joint Modeling to Estimate
the Treatment Eﬀect on a Longitudinal Surrogate with Application to Chronic Kidney Disease Trials," Biometrics, 81(3): ujaf104, (<https://doi.org/10.1093/biomtc/ujaf104>).

If you are viewing the html version of this tutorial, you can find the R markdown file here: <https://github.com/laylaparast/longsurr>. It is the file called longsurr_tutorial.Rmd.

### Evaluating a Longitudinal Surrogate Marker: Agniel and Parast (2021) 

Let's first take a look at the example dataset included in the package, **full_data**. The dataset has 10100 observations and contains a uniqie person ID, **id**, the treatment group, **a**,  times that the surrogate marker was measured, **tt**, the longitudinal surrogate **x**, and the primary outcome **y**. 

```{r}
data(full_data)
names(full_data)
head(full_data)
```

Our goal is to evaluate the proportion of the treatment effect (PTE) on the primary outcome that is explained by the treatment effect on the longitudinal surrogate. Let's take a look at the estimated PTE using the Agniel and Parast (2021) estimation methods with the example data described above. The function **estimate_surrogate_value** provides the estimated treatment effect on the primary outcome, **Deltahat**, the residual treatment effect, **Deltahat_S** and the PTE, denoted as **R**. The estimator **type** can be either 'gam', 'linear', or 'kernel', we first use linear below with **var=FALSE**. We use **dplyr** to reshape our dataset. 

```{r results = "hide", message=FALSE}
library(dplyr)
```

```{r}
wide_ds <- full_data %>% 
dplyr::select(id, a, tt, x, y) %>%
tidyr::spread(tt, x) 

wide_ds_0 <- wide_ds %>% filter(a == 0)
wide_ds_1 <- wide_ds %>% filter(a == 1)
X_t <- wide_ds_1 %>% dplyr::select(`-1`:`1`) %>% as.matrix
y_t <- wide_ds_1 %>% pull(y)
X_c <- wide_ds_0 %>% dplyr::select(`-1`:`1`) %>% as.matrix
y_c <- wide_ds_0 %>% pull(y)

estimate_surrogate_value(y_t = y_t, y_c = y_c, X_t = X_t, X_c = X_c, 
method = 'linear', var = FALSE)
```

Next, we use "gam" estimation and set **var=TRUE** to get variance estimates and confidence intervals.

```{r}

estimate_surrogate_value(y_t = y_t, y_c = y_c, X_t = X_t, X_c = X_c, 
method = 'gam', var = TRUE)

```

### Estimating the Treatment Effect on a Longitudinal Surrogate Marker

Let's now take a look at the dataset **data_sjm** which we will use to illustrate the semiparametric joint modeling methods. Here our aim is to estimate the treatment effect on a lognitudinal surrogate in the presence of a censored outcome.  This dataset contains 200 observations containing the event indicator **delta**, the time that the surrogate marker was measured for each observation, **optT**, the surrogate marker measurements over time for each observation, **Y**, the observed event or censoring time for each observation, **Time**, and the treatment indicator for each observation with 1 for treated and 0 for control, **Treatment**.

```{r}
data(data_sjm)
names(data_sjm)
```

This packages offers to methods this treatment effect assuming either a linear model or a nonlinear model for the surrogate trajectory. We first illustrate linear below with **var=TRUE**. 

```{r}
sjm_linear_estimate(X=data_sjm$Treatment, Time = data_sjm$Time, 
Delta = data_sjm$delta, obsT = data_sjm$obsT, Y = data_sjm$Y, n.resample=5, var=TRUE)

```
 The function returns the point estimates for the hazard ratio, the treatment effect on the baseline surrogate measurement, and the treatment effect on the surrogate trajectory, **est**. Since **var=TRUE**, the function also returns the corresponding standard error estimates, and lower and upper bounds of the 95\% confidence intervals, respectively, **SE, CI_lower, CI_upper**. Next, we use the nonlinear estimator with   **var=FALSE**. For the nonlinear function, the user must additionally specify a **gap_time** used to create the grid for the trakectory estimates. 
 
```{r}
sjm_nl_estimate(X=data_sjm$Treatment, Time = data_sjm$Time, 
Delta = data_sjm$delta, obsT = data_sjm$obsT, Y = data_sjm$Y, var=FALSE, gap_time=0.2)

```
 The function returns the hazard ratio estimate, **est**, and the estimated treatment effect on the surrogate trajectory, **est_t**, estimates on the grid points **t_grid**. If **var=TRUE**, similar to the linear function, corresponding standard errors and confidence bounds are returned. 

That's all for now. Please email <parast@austin.utexas.edu> if you have questions.


------------------------------------------------------------------------